PRISM protocol v0.3

### 2021-10-12 prism-backend/research/article.tex
Updated description of the connector protocol and its security properties.

\documentclass[10pt,a4paper]{article}
\usepackage[margin=1in]{geometry}

\newcommand{\todo}[1]{{\colorbox{red}{\bf TODO:}\textcolor{red}{#1}}}

@@

In a nutshell, the proposed protocol is composed essentially of two phases. In
the first phase, the issuer shares a QR code with the holder. This QR code
contains the issuer's public key (or, maybe, its DID), a random value, and a
session identifier. This exchange takes place through a secure (i.e.,
confidential and authenticated) channel, such as physical exchange. Probably
immediately, although also possibly at a later point in time, the prospective
holder sends back her own public key (or, alternatively, her DID) and the
session identifier, all authenticated with a MAC, where we use the random value
in the QR code as secret key. The issuer uses the session identifier to locate
the corresponding session and, now that it received in an authenticated manner
the public key of the holder, can leverage hybrid encryption techniques to
establish an end-to-end encrypted session, e.g., using ECIES \cite{abr01}.
Ensuring end-to-end encryption and authentication is important at this point, as
all the information exchanged between issuer and holder is routed through the
connector: an entity or set of entities who ensure that asynchronous
communication can take place between issuer and holder. Thus, ECIES is crucial
to ensure the secrecy and authenticity of the information, end-to-end.

We have modelled this protocol using Tamarin%
\footnote{https://tamarin-prover.github.io/ (Last access, October 5th, 2021.)}

>>>>

\documentclass[10pt,a4paper]{article}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage{graphicx}

\newcommand{\figref}[1]{Fig. \ref{#1}}
\newcommand{\tabref}[1]{Table \ref{#1}}
\newcommand{\lstref}[1]{Listing \ref{#1}}
\newcommand{\secref}[1]{Section \ref{#1}}
\newcommand{\appref}[1]{Appendix \ref{#1}}
\newcommand{\todo}[1]{{\colorbox{red}{\bf TODO:}\textcolor{red}{#1}}}

@@

In a nutshell, the proposed protocol is composed essentially of two phases. In
the first phase, the issuer shares a QR code with the holder. This QR code
contains the issuer's public key (or, maybe, its DID), a random value, and a
session identifier. This exchange takes place through a secure (i.e.,
confidential and authenticated) channel, such as physical exchange. Probably
immediately, although also possibly at a later point in time, the prospective
holder sends back her own public key (or, alternatively, her DID) and the
session identifier, all authenticated with a MAC, where we use the random value
in the QR code as secret key. The issuer uses the session identifier to locate
the corresponding session and, now that it received in an authenticated manner
the public key of the holder, can leverage hybrid encryption techniques to
establish an end-to-end encrypted session, e.g., using ECIES \cite{abr01}.
Ensuring end-to-end encryption and authentication is important at this point, as
all the information exchanged between issuer and holder is routed through the
connector: an entity or set of entities who ensure that asynchronous
communication can take place between issuer and holder. Thus, ECIES is crucial
to ensure the secrecy of the information, end-to-end. Since the holder needs to
have certainty that the data it receives comes from the issuer (and ECIES does
not authenticate the sender), we have the issuer digitally sign the data it
sends through the ECIES-secured channel. Note that we could involve the random
value shared via the QR code for this. But that would force us to re-execute the
first phase of the protocol even when the QR code is compromised after having
completed the first phase of the protocol. Digital signatures using the issuer's
DID prevent this.

While the protocol is simple, and is based on common cryptographic constructions,
it is easy to get such protocols wrong, as pointed out in numerous occasions
\cite{davi01, kraw01}. Hence, we have used Tamarin%
\footnote{https://tamarin-prover.github.io/ (Last access, October 5th, 2021.)}
to model our protocol\footnote{Source code for the model available here
  \url{https://github.com/input-output-hk/atala-prism/blob/research-tamarin-connector/prism-backend/research/formalizations/connector.spthy}
  \todo{Update to final URL.}}, which is schematized in \figref{fig:e2econnector}

\begin{figure}[ht!]
  \centering
  \includegraphics[scale=0.4]{figures/connector.png}
  \caption{Sequence diagram for the E2EE protocol through the connector.}
  \label{fig:e2econnector}%
\end{figure}

The security properties that we prove are as follows (we refer to the actual
model for the formal definition):

\begin{description}
\item[Code secrecy.] Codes exchanged via a QR code are always confidential.
  Note that this is trivial, as it is exchanged via a secure channel, and
  the holder never sends it back in the clear.
\item[Code injectivity.] Every QR code that is received by a holder, allegedly
  sent by an issuer, was sent before by that issuer. Again, this is trivial due
  to the secure channel used to exchange the QR code.
\item[Holder's public key authenticity.] Every message received by an issuer
  in step 3, containing a public key for key-exchange usage, that was
  allegedly sent by a holder, was indeed previously sent by that holder.
  This is ensured through the message authentication code (MAC) that
  leverages the random QR code exchanged via the secure channel, which is
  only known to the issuer and the holder.
\item[Holder's public key injectivity.] For every session in which an
  issuer receives in step 3 a message, allegedly sent by a holder in
  step 2, containing the holder's public key, it was that holder, in
  the same session, who sent the public key in step 2.  
\item[Credential secrecy.] This property has two variants: it ensures
  secrecy for credentials sent by an honest issuer to a holder that was honest
  at the moment of sending the credential, and is not compromised afterwards;
  and it also ensures secrecy for credentials received by an honest holder,
  allegedly from an honest issuer, if both holder and issuer were honest when
  the credential is received, and they are not corrupted afterwards. Altogether
  with the credential authenticity property shown next, it ensures that only
  (honest) holder and issuer learn the contents of a credential.
\item[Credential authenticity.] All messages containing a credential, received
  by a holder at step 5, where indeed sent by an issuer and addressed to that
  same holder, as long as both issuer and holder keys are not compromised.
\item[Credential non-injectivity.] All messages containing a credential, received
  by a holder at step 5, were sent by an issuer and addressed to that holder
  at step 4. Note, however, that both messages in steps 4 and 5 dont necessarily
  pertain to the  same ``session'' (in instantiations where parallel sessions
  may occur). However, this does not pose a risk, as in any case, the messaged
  is sent by the issuer, and received by the intended holder.
\end{description}

Concerning the technical details of the protocol, note that we are combining
signature and encryption (plus, the MAC-based authentication natively provided
by ECIES). This places us in the context put forward in \cite{davi01}, that
discusses some good practices when combining such primitives. In a nutshell,
we follow the sign-then-encrypt approach\footnote{What we call here ``sign-then-
  encrypt'' is called ``Sign \& Encrypt'' in \cite{davi01}. However, we find
  more explicit the naming convention of \cite{kraw01}, which is more consistent
  when referring to the order in which the primitives are used -- order that, in
  these cases, is crucial.}, combined with including the public
keys of both holder and issuer into the signed (and then encrypted) data.
This intertwines the inner layer (the signature) with the outer layer (the
encryption), ensuring that any alteration on any of them will be detected.
We note that there are other options that, a priori, are equally secure -- e.g.,
encrypt-then-sign, including also the sender's public key in the encrypted data
(but that would require separately signing a (hash of) the data, to ensure
non-repudiation). It is worth noting that ECIES follows the
encrypt-then-authenticate approach, which is proven to realize secure channels
in \cite{kraw01}, as long as the underlying encryption scheme is IND-CPA secure,
and the underlying MAC function is secure.
